---
title: Which optimization algorithm to choose?
author: Marie Laure Delignette Muller, Christophe Dutang
date: '`r Sys.Date()`'
output:
  html_vignette:
    toc: yes
    number_sections: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Optimization algorithm} 
  %!\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---
```{r, message=FALSE}
require(fitdistrplus)
require(actuar)
require(knitr)
```



# Quick overview of main optimization methods

For a good introduction to books such as Nocedal \& Wright (2006) or Bonnans, Gilbert, Lemarechal \& Sagastiz\'abal (2006). We consider
the problem $\min_x f(x)$ for $x\in\mathbb{R}^n$.

## Derivative-free optimization methods
The Nelder-Mead method is one of the most well known derivative-free methods
that use only values of $f$ to search for the minimum.
It consists in building a simplex of $n+1$ points and moving/shrinking
this simplex into the good direction.
\begin{enumerate}
\item set initial points $x_1, \dots, x_{n+1}$.
\item order points such that $f(x_1)\leq f(x_2)\leq\dots\leq f(x_n)$.
\item compute $x_o$ as the centroid of $x_1, \dots, x_{n}$. 
\item Reflection: 
  \begin{itemize}
  \item compute the reflected point $x_r = x_o + \alpha(x_o-x_{n+1})$
  \item \textbf{if} $f(x_1)\leq f(x_r)<f(x_n)$, 
  then replace $x_{n+1}$ by $x_r$,
  go to step 2.
  \item \textbf{else} go step 5.
  \end{itemize}
\item Expansion:   
  \begin{itemize}
  \item \textbf{if} $f(x_r)<f(x_1)$, then compute the expansion point
    $x_e= x_o+\gamma(x_o-x_{n+1})$.
    \begin{itemize}
    \item[] \textbf{if} $f(x_e) <f(x_r)$, then replace $x_{n+1}$ by $x_e$,
  go to step 2.
    \item[] \textbf{else} $x_{n+1}$ by $x_r$, go to step 2.
    \end{itemize}  
  \item \textbf{else} go to step 6.
  \end{itemize}
\item Contraction:  
  \begin{itemize}
  \item compute the contracted point $x_c = x_o + \beta(x_o-x_{n+1})$
  \item \textbf{if} $f(x_c)<f(x_{n+1})$, 
  then replace $x_{n+1}$ by $x_c$,
  go to step 2.
  \item \textbf{else} go step 7.
  \end{itemize}
\item Reduction:  
  \begin{itemize}
  \item for $i=2,\dots, n+1$, compute $x_i = x_1+\sigma(x_i-x_{1})$
  \end{itemize}
\end{enumerate}
The Nelder-Mead method is available in `optim`.
By default, in `optim`, $\alpha=1$, $\beta=1/2$, $\gamma=2$ and $\sigma=1/2$.

## Hessian-free optimization methods

For smooth non-linear function, the following method is generally used:
a local method combined with line search work on the scheme $x_{k+1} =x_k + t_k d_{k}$, where the local method will specify the direction $d_k$ and the line search will specify the step size $t_k \in \mathbb{R}$.	

### Computing the direction $d_k$
A desirable property for $d_k$ is that $d_k$ ensures a descent $f(x_{k+1}) < f(x_{k})$. 
Newton methods are such that $d_k$ minimizes a local quadratic approximation of $f$ based on a Taylor expansion, that is  $q_f(d) = f(x_k) + g(x_k)^Td +\frac{1}{2} d^T H(x_k) d$ where $g$ denotes the gradient and $H$ denotes the Hessian.

The \emph{exact Newton} consists in using the exact solution of local minimization problem $d_k = - H(x_k)^{-1} g(x_k)$.	
In practice, other methods are preferred (at least to ensure positive definiteness).
The \emph{quasi-Newton} method approximates the Hessian by a matrix $H_k$ as a function of $H_{k-1}$, $x_k$, $f(x_k)$ and then $d_k$ solves the system $H_k d = -  g(x_k)$. 
Some implementation may also directly approximate the inverse of the Hessian $W_k$ in order to compute $d_k = -W_k g(x_k)$. Using the Sherman-Morrison-Woodbury formula, we can switch between $W_k$ and $H_k$.

To determine $W_k$, first it must verify the secant equation $H_k y_k =s_k$ or $y_k=W_k s_k$ where $y_k = g_{k+1}-g_k$ and $s_k=x_{k+1}-x_k$. To define the $n(n-1)$ terms, we generally impose a symmetry and a minimum distance conditions. We say we have a rank 2 update if  $H_k = H_{k-1} + a u u^T + b v v^T$ and a rank 1 update if  $H_k = H_{k-1} + a u u^T $. Rank $n$ update is justified by the spectral decomposition theorem.


There are two rank-2 updates which are symmetric and preserve positive definiteness 
\begin{itemize}
\item DFP \emph{Davidon-Fletcher-Powell} minimizes $\min || H - H_k ||_F$ such that $H=H^T$:
$$ 
H_{k+1} = \left (I-\frac {y_k s_k^T} {y_k^T s_k} \right ) H_k \left (I-\frac {s_k y_k^T} {y_k^T s_k} \right )+\frac{y_k y_k^T} {y_k^T s_k}  
\Leftrightarrow
W_{k+1} = W_k +  \frac{s_k s_k^T}{y_k^{T} s_k} - \frac {W_k y_k y_k^T W_k^T} {y_k^T W_k y_k} .
$$
\item BFGS \emph{Broyden-Fletcher-Goldfarb-Shanno} minimizes $\min || W - W_k ||_F$ such that $W=W^T$:
$$
H_{k+1} = H_k - \frac{ H_k y_k y_k^T H_k }{ y_k^T H_k y_k }  + \frac{ s_k s_k^T }{ y_k^T s_k }
\Leftrightarrow
W_{k+1} = \left (I-\frac {y_k s_k^T} {y_k^T s_k} \right )^T W_k \left (I-\frac { y_k s_k^T} {y_k^T s_k} \right )+\frac{s_k s_k^T} {y_k^T s_k} .
$$
\end{itemize}
In `R`, the so-called BFGS scheme is implemented in `optim`.
The DFP method is not available?

Another possible method (which is initially arised from quadratic problems)
is the nonlinear conjugate gradients.
This consists in computing directions $(d_0, \dots, d_k)$ that are conjugate
with respect to a matrix close to the true Hessian $H(x_k)$.
Directions are computed iteratively by $d_k = -g(x_k) + \beta_k d_{k-1}$ for $k>1$, once initiated by $d_1 = -g(x_1)$. 
$\beta_k$ are updated according a scheme: 
\begin{itemize}
\item $\beta_k = \frac{ g_k^T g_k}{g_{k-1}^T g_{k-1} }$: Fletcher-Reeves update,
\item $\beta_k = \frac{ g_k^T (g_k-g_{k-1} )}{g_{k-1}^T g_{k-1}}$: Polak-Ribi\`ere update.
\end{itemize}
There exists also three-term formula for computing direction 
$d_k = -g(x_k) + \beta_k d_{k-1}+\gamma_{k} d_t$ for $t<k$.
A possible scheme is the Beale-Sorenson update defined as
 $\beta_k = \frac{ g_k^T (g_k-g_{k-1} )}{d^T_{k-1}(g_{k}- g_{k-1})}$ 
 and $\gamma_k = \frac{ g_k^T (g_{t+1}-g_{t} )}{d^T_{t}(g_{t+1}- g_{t})}$ if $k>t+1$ otherwise $\gamma_k=0$ if $k=t$.
See Yuan (2006) for other well-known schemes such as Hestenses-Stiefel, Dixon or Conjugate-Descent.
This three updates of the (non-linear) conjugate gradient are available in `optim`.


### Computing the stepsize $t_k$

Let $\phi_k(t) = f(x_k + t d_k)$ for a given direction/iterate $(d_k, x_k)$. 
We need to find conditions to find a satisfactory stepsize $t_k$. In literature, we consider the  descent condition: $\phi_k'(0) < 0$
and the Armijo condition: $\phi_k(t) \leq \phi_k(0) + t c_1 \phi_k'(0)$ ensures a decrease of $f$.
Nocedal \& Wright (2006) presents a backtracking (or geometric) approach satisfying the Armijo condition and minimal condition, i.e. Goldstein and Price.
\begin{itemize}
\item set $t_{k,0}$ e.g. 1, $0 < \alpha < 1$,
\item \textbf{Repeat} until Armijo satisfied,
	\begin{itemize}
	\item[] $t_{k,i+1} =  \alpha \times t_{k,i}$.
	\end{itemize} 
\item \textbf{end Repeat}
\end{itemize}
This backtracking linesearch is available in `optim`.


## Benchmark

To simplify the benchmark of optimization methods, we create a `fitbench` function that computes
the desired estimation method for all optimization methods. This function is currently not exported.
```{r, echo=TRUE, eval=FALSE}
fitbench <- function(data, distr, method, grad=NULL, control=list(trace=0, REPORT=1, maxit=1000), lower=-Inf, upper=+Inf, ...) 
```
```{r, echo=FALSE}
fitbench <- fitdistrplus:::fitbench
```



# Log-likelihood function and its gradient for beta distribution

## Theoretical value
The density of the beta distribution is given by 
$$
f(x; \delta_1,\delta_2) = \frac{x^{\delta_1-1}(1-x)^{\delta_2-1}}{\beta(\delta_1,\delta_2)},
$$
where $\beta$ denotes the beta function, see the NIST Handbook of mathematical functions.
We recall that $\beta(a,b)=\Gamma(a)\Gamma(b)/\Gamma(a+b)$.
There the log-likelihood for a set of observations $(x_1,\dots,x_n)$ is
$$
\log L(\delta_1,\delta_2) = (\delta_1-1)\sum_{i=1}^n\log(x_i)+ (\delta_2-1)\sum_{i=1}^n\log(1-x_i)+ n \log(\beta(\delta_1,\delta_2))
$$
The gradient with respect to $a$ and $b$ is 
$$
\nabla \log L(\delta_1,\delta_2) = 
\left(\begin{matrix}
\sum\limits_{i=1}^n\ln(x_i) - n\psi(\delta_1)+n\psi( \delta_1+\delta_2)  \\
\sum\limits_{i=1}^n\ln(1-x_i)- n\psi(\delta_2)+n\psi( \delta_1+\delta_2)
\end{matrix}\right),
$$
where $\psi(x)=\Gamma'(x)/\Gamma(x)$ is the digamma function, 
see the NIST Handbook of mathematical functions.

## `R` implementation
As in the `fitdistrplus` package, we minimize the opposite of the log-likelihood: we implement the opposite of the gradient in `grlnL`.
```{r}
lnL <- function(par, fix.arg, obs, ddistnam) 
  fitdistrplus:::loglikelihood(par, fix.arg, obs, ddistnam) 
grlnlbeta <- fitdistrplus:::grlnlbeta
```




# Numerical application to beta distribution


## Random generation of a sample

```{r, fig.height=5, fig.width=5}
#(1) beta distribution
n <- 200
set.seed(12345)
x <- rbeta(n, 3, 3/4)
grlnlbeta(c(3, 4), x) #test
hist(x, prob=TRUE)
lines(density(x), col="red")
curve(dbeta(x, 3, 3/4), col="green", add=TRUE)
legend("topleft", lty=1, col=c("red","green"), leg=c("empirical", "theoretical"))
```

## Fit Beta distribution

Define control parameters.
```{r}
ctr <- list(trace=0, REPORT=1, maxit=1000)
```
Call `mledist` with the default optimization function 
(`optim` implemented in `stats` package) 
with and without the gradient for the different optimization methods.
```{r}
unconstropt <- fitbench(x, "beta", "mle", grad=grlnlbeta)
```
In the case of constrained optimization, `mledist` permits the direct use 
of `constrOptim` function (still implemented in `stats` package)
that allow linear inequality constraints by using a logarithmic barrier.

Use a exp/log transformation of the shape parameters $\delta_1$ and $\delta_2$ 
to ensure that the shape parameters are strictly positive.
```{r}
dbeta2 <- function(x, shape1, shape2, log)
  dbeta(x, exp(shape1), exp(shape2), log=log)
#take the log of the starting values
startarg <- lapply(fitdistrplus:::start.arg.default(x, "beta"), log)
#redefine the gradient for the new parametrization
grbetaexp <- function(par, obs, ...) 
    grlnlbeta(exp(par), obs) * exp(par)
    

expopt <- fitbench(x, distr="beta2", method="mle", grad=grbetaexp, start=startarg) 
#get back to original parametrization
expopt[c("fitted shape1", "fitted shape2"), ] <- exp(expopt[c("fitted shape1", "fitted shape2"), ])
```
Then we extract the values of the fitted parameters, the value of the 
corresponding log-likelihood and the number of counts to the function
to minimize and its gradient (whether it is the theoretical gradient or the
numerically approximated one).



## Results of numerical investigation
Results are displayed in the following tables.

```{r, results='asis', echo=FALSE}
kable(unconstropt)
```

```{r, results='asis', echo=FALSE}
kable(expopt)
```


Using `llsurface`, we plot the log-likehood surface around the true value (green) and the fitted parameters (red).
```{r, fig.width=5, fig.height=5}
llsurface(min.arg=c(0.1, 0.1), max.arg=c(7, 3), 
          plot.arg=c("shape1", "shape2"), nlev=25,
          plot.np=50, data=x, distr="beta")
points(unconstropt[1,"BFGS"], unconstropt[2,"BFGS"], pch="+", col="red")
points(3, 3/4, pch="x", col="green")
```

We can simulate bootstrap replicates using the `bootdist` function.
```{r, fig.width=5, fig.height=5}
b1 <- bootdist(fitdist(x, "beta", method="mle", optim.method="BFGS"), niter=100, parallel="snow", ncpus=2)
summary(b1)
plot(b1, enhance=TRUE, trueval=c(3, 3/4))
```



## Conclusion

Based on the beta example, we observe that all methods converge to the same
point. This is rassuring.  
```{r}
unconstropt[1:2,"BFGS"]
```
However, the number of function evaluation (and gradient evaluation) is
very different from a method to another.
Futhermore, specifying the true gradient of the log-likelihood does not
help at all the fitting procedure and generally slows down the convergence.
The best methods only evaluates 17 times the log-likelihood.
```{r}
resall <- cbind(unconstropt, expopt)
cmin <- min(resall["func. eval. nb.", ], na.rm=TRUE)
colnames(resall)[resall["func. eval. nb.", ] == cmin]
```
That is BFGS with the exponential transformation of the parameters.
Since the exponential function is differentiable, the asymptotic properties are
still preserved (by the Delta method) but for finite-sample this may produce a small bias.
The second-best method is the BFGS with neither with gradient, neither with constaint, 
nor with exponential transform.
The difference between BFGS with and without exponential transformation 
in terms of fitted values are neglible.





