---
title: "Explanatory note for NNLS"
author: "Christophe Dutang"
date: "5/15/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The NNLS problem

Consider the non-negative least square problem with a sum constraint.
$$
\min_{x} ||A x - b||^2, \text{ such that } <x,1>=c, \forall i, x_i\geq0,
$$
where $A\in\mathbb R^{m\times n}$, $b\in\mathbb R^{m}$ and $c\in\mathbb R^{n}$. 

# Reformulation by integrating the sum constraint
Let define the 
$$
\theta = 
\left(\begin{matrix} x_1\\ \vdots\\ x_{n-1}
\end{matrix}\right)
\Rightarrow
x_\theta=
\left(\begin{matrix} x_1\\ \vdots\\ x_{n-1} \\ c-<1,\theta> 
\end{matrix}\right)
=\left(\begin{matrix} \theta \\ c-<1,\theta>
\end{matrix}\right).
$$
The product matrix at that point is
$$
A x_\theta  = 
\left(\begin{matrix} 
A_{1,1} & \dots & A_{1,n-1} & A_{1,n} \\ &\vdots \\ 
A_{m,1} & \dots & A_{m,n-1} & A_{m,n}
\end{matrix}\right)
\left(\begin{matrix} x_1\\ \vdots\\ x_{n-1} \\ c-<1,\theta> 
\end{matrix}\right)
=
\left(\begin{matrix} 
A_{1,1}x_1 + \dots + A_{1,n-1}x_{n-1} + A_{1,n}(c-<1,\theta>) \\ \vdots \\ 
A_{m,1}x_1 + \dots + A_{m,n-1} + A_{m,n}(c-<1,\theta>)
\end{matrix}\right)
$$
$$
=
A_{,\leq n-1}\theta + A_{,n} (c-<1,\theta>)
$$
where $A=(A_{,\leq n-1}, A_{,n})$.

So we define the objective function
$$
F_c(\theta) = ||A x_\theta - b||^2
= (A_{,\leq n-1}\theta + A_{,n} (c-<1,\theta>) - b)^T(A_{,\leq n-1}\theta + A_{,n} (c-<1,\theta>) - b)
$$

Let us compute the derivative for $i\leq n-1$
$$
\frac{\partial (A x_\theta - b)}{\partial \theta_i}
= 
\frac{\partial (A_{,\leq n-1}\theta ))}{\partial \theta_i}
+
\frac{\partial (A_{,n} (c-<1,\theta>))}{\partial \theta_i}
= A_{,i} - A_{,n}.
$$
The derivative for $F_c$ is $i\leq n-1$
$$
\frac{\partial F_c(\theta)}{\partial \theta_i}
=2(A_{,i} - A_{,n})^T (A_{,\leq n-1}\theta + A_{,n} (c-<1,\theta>) - b).
$$
So the gradient for $F_c$ is
$$
\nabla F_c(\theta)
=2(A_{,\leq n-1} - A_{,n})^T (A_{,\leq n-1}\theta + A_{,n} (c-<1,\theta>) - b).
$$

# Numerical check

Quick check on R
```{r}
Mc <- function(theta, a, b, sumtotal)
{
  x <- c(theta, sumtotal-sum(theta))
  a %*% x - b
}
gradMc <- function(i, a, b, sumtotal)
{
  a[, i] - a[, NCOL(a)]
}
  
Fc <- function(theta, a, b, sumtotal)
{
  x <- c(theta, sumtotal-sum(theta))
  sum((a %*% x - b)^2)
}
gradFc <- function(theta, i, a, b, sumtotal)
{
  x <- c(theta, sumtotal-sum(theta))
  diffa <- a[, i] - a[, NCOL(a)]
  y <- a %*% x - b
  2*crossprod(diffa, y)
}
  
fullgradFc <- function(theta, a, b, sumtotal)
{
  x <- c(theta, sumtotal-sum(theta))
  sub_a <- a[, 1:(NCOL(a)-1)]
  diffa <- sub_a - a[, NCOL(a)]
  y <- a %*% x - b
  2*crossprod(diffa, y)
}
  

a <- matrix(1:12, 4, 3)
b <- 1:4/4

theta1 <- 1:2
theta2 <- theta1+c(0,1e-3)
(Mc(theta1, a, b, 1) - Mc(theta2, a, b, 1))/-1e-3
  
gradMc(2, a, b, 1)

gradFc(theta1, 2, a, b, 1)
(Fc(theta1, a, b, 1) - Fc(theta2, a, b, 1))/-1e-3

fullgradFc(theta1, a, b, 1)
sapply(1:2, function(i) gradFc(theta1, i, a, b, 1))
```



